Abstractions for Multi-Sorted Substitutions

### 2024-03-18 paper/paper.tex
Update.

  Code of the framework is displayed in gray boxes.
  Code of examples is written in yellow boxes.
  Code of the System F formalization is written without boxes. The
  latter is the only code a user of our framework has to write.

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in our supplementary material,
  which includes the simplified framework (368 lines of code) and the
  System F formalization (221 lines of code).
  The full framework including the case studies is available on Github:
  \textcolor{blue}{\url{https://github.com/m0rphism/kitty}}
  % The full framework is available online.\footnote{See supplementary material. In the final version, this text will be replaced by a link to github.}

@@

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  % \item
  %   an adaptation of McBride's
  %   kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  %   for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and type contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  We identify two drawbacks with this style of syntax:
  \begin{enumerate}
  \item
    the syntactic categories (\ACode{\ACon{Kind}},
    \ACode{\ACon{Type}}, and \ACode{\ACon{Expr}})
    have different types, which makes it difficult to treat them uniformly; and
  \item the different sorts of variables are modeled separately, which requires to define not just
    type-in-type and expression-in-expression substitution, but also type-in-expression substitution.
    Consequently, interaction lemmas between those substitutions are required.
    % In general, this leads to a combinatory explosion of
    % substitutions, e.g.\ adding kind variables would lead to a total of 6
    % different substitutions and corresponding interaction lemmas.
  \end{enumerate}
  To avoid these drawbacks, we instead use a multi-sorted syntax.

@@

  \begin{figure}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
    \label{fig:map-ops}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions.
  The definitions are included directly in the record module of the
  Kit, so they are implicitly parameterized over a kit.
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \& x} applies a map to a variable.
  The operation \ACode{ϕ ↑ s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{⦅ x/t ⦆} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\AField{weaken s}} map increases all variables by one.
  The \ACode{ϕ₁ $\sim$ ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{$\sim$-ext}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{id}.
  The lemma \ACode{id↑-id} states that a lifted identity map is again an identity map.

  To make it easier to talk about a specific kit, we introduce the following notations:
  % We also specify notation 
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  % We also specify notation, which makes it less cumbersome
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  \begin{itemize}
  \item
     we write \ACode{(S \ADef{∋/⊢[} K \ADef{]} s)} for the \ACode{(S ∋/⊢ s)}
     of some specific \ACode{\ACon{Kit}} \ACode{K}; and
   \item
     we write \ACode{(S₁ \ADef{–[} K \ADef{]→} S₂)} for the \ACode{(S₁ \ADef{→ₖ} S₂)}
     of some specific \ACode{\ACon{Kit}} \ACode{K}.
  \end{itemize}

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  \noindent The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

@@

  First, we define the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}}. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

@@

  A compose kit \ACode{\ACon{CKit}~K₁~K₂~K₁⊔K₂}
  describes the operations necessary for definining the composition of a
  map from \ACode{K₁} with a map from \ACode{K₂} that results in a map from
  \ACode{K₁⊔K₂}:
  \LibCode\KCKit
  \noindent The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency\cite{DBLP:conf/esop/Jones00} and is determined
  by the choice of \ACode{K₁} and \ACode{K₂}.
  The fields of a compose kit have the following meaning:
  \begin{itemize}
  \item
    The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
    \ACode{x/t} (according to \ACode{K₁}) and a renaming or substitution
    \ACode{ϕ} (according to \ACode{K₂}) and applies \ACode{ϕ} to \ACode{x/t} resulting in
    a variable or term (according to \ACode{K₁⊔K₂}).
    From this operation we derive map composition \ACode{\ADef{\_·ₖ\_}} as
    shown in the previous subsection.
    % We use the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}, if we need to
    % make it explicit that we are using the composition of some
    % specific \ACode{\ACon{CKit}} \ACode{C}.
  \item 
    The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
    \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
    subsequent lemmas to make use of the lemmas that we have already proved
    for \ACode{\AField{\_⋯\_}}.
  \item 
    The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
    weakening is the same as weakening first and then lifting the map over
    the variable introduced by the weakening.
    From this lemma, we can derive that lifting distributes over composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}

@@

    % A weakening followed by a singleton substitution cancel each other out:
    % \LibCode\KWeakenCancelsSingle
    % % \LibCode\KWeakenCancelsSingleProof
    % By applying \ACode{\AField{fusion}} and \ACode{\AField{⋯-id}} to
    % this lemma, we get the same statement for multiple map applications:
    % \LibCode\KWeakenCancelsSingleTraverse
    % % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingleTraverse
    % A singleton map can be swapped with any map \ACode{ϕ}:
    % \LibCode\KDistLiftSingle
    % % \LibCode\KDistLiftSingleProof
    % By applying \ACode{\AField{fusion}} to this lemma, we get the
    % same statement for multiple map applications:
    % \LibCode\KDistLiftSingleTraverse
    % % \LibCode\KDistLiftSingleTraverseProof
  \end{itemize}
  Similarly, as it was the case for the \ACode{\ADef{Kit}} and \ACode{\ADef{Traversal}}
  structures, the idea is that we instantiate the \ACode{\ADef{CTraversal}} for
  our object language, and in return the framework defines the concrete \ACode{\ADef{CKit}}
  instances for us. Hence, we define them in the record module of \ACode{\ADef{CTraversal}}:
  \LibCode\KCKitInstances
  \noindent
  \ACode{\ADef{Cᵣ}} is the compose kit between a renaming and another kit \ACode{K}.
  \ACode{\ADef{Cₛ}} is the compose kit between a substitution and another kit \ACode{K},
  and requires that we already know how to compose a \ACode{K}-map with a renaming.
  The following verifies that \ACode{\ADef{Cᵣ}} and \ACode{\ADef{Cₛ}} indeed get us all four compositions:

@@

  \AppCode\FPreserve
  The type of \ACode{\ADef{\_⊢⋯\_}} is the same as in the record definition and hence omitted.
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\AField{\_⋯\_}}. Where \ACode{\AField{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, we now use their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use
    \ACode{\AField{⋯-↑-wk}} to move the map under the weakening.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use \ACode{\AField{dist-↑-⦅⦆}} to move the map
    under the singleton substitution.
  \end{itemize}
  % Finally, we instantiate the \ACode{\ACon{TTraversal}} structure:
  % \AppCode\FTTraversal

@@

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    works out of the box for both deriving untyped substitution
    and instantiating the \ACode{\ACon{TTraversal}}.
    In the proof of confluence, the \ACode{\ACon{CKit}} abstraction
    allowed us to unify lemmas about the reduction of renamings and
    substitutions.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution-preserves-typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we instead used an encoding that allows us to use the
    the \ACode{\ACon{TKit}} abstraction directly.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accesses by the user.
    With this encoding, substitution-preserves-typing is generally true
    again: replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t}, results in a term that is still
    well-typed, but in a context with an unsatisfiable constraint \ACode{t' <: t}.
  \item
    Object languages with pattern matching can be encoded by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself:
    \begin{quote}
    \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    \end{quote}
    This allows the clause of a \ACode{\ACon{match}} expression to
    read the bound variables from the pattern sort and make them
    available as free variables in its body expression.
  \end{itemize}

@@

  \subsection{Variable Binding}
  There is a plethora of different methods for representing
  variable binders in the literature:
  de Bruijn\cite{scholar:de1972lambda},
  co de Bruijn\cite{DBLP:journals/corr/abs-1807-04085},
  locally nameless\cite{DBLP:journals/jar/Chargueraud12},
  locally named\cite{DBLP:conf/tlca/McKinnaP93},
  higher order abstract syntax\cite{DBLP:conf/pldi/PfenningE88} and
  its parametric variant\cite{DBLP:conf/icfp/Chlipala08},
  nominal logic\cite{DBLP:journals/tcs/UrbanPG04},
  shifted names\cite{scholar:dolansyntax},
  nameless painless\cite{DBLP:conf/icfp/Pouillard11}, and
  scope graphs\cite{DBLP:journals/pacmpl/AntwerpenPRV18}.
  Many of these representations have been studied in solutions to the POPLMark
  challenges\cite{DBLP:conf/tphol/AydemirBFFPSVWWZ05, DBLP:journals/jfp/AbelAHPMSS19}.

@@

  With only a slight modification to their framework, we can instantiate it
  for multi-sorted syntax, enabling the definition of polymorphic languages.
  However, as the intrinsic typing is then
  used to ensure syntactic sorts, the semantic abstractions talk about untyped
  terms, so typing and type preservation lemmas have to be modeled entirely manually.
  We believe it would be worthwhile to explore how their semantics
  abstractions can be lifted to typing relations similar to how our typing kits lift
  regular kits from terms to typing relations.

@@

  While intrinsic typing allows to unify definitions with their type
  preservation proofs, extrinsic typing allows to unify
  substitutions across different syntactic categories, as we have demonstrated.
  We believe this makes extrinsic typing more suitable for polymorphic languages:
  with extrinsic typing, the definitions and lemmas for untyped
  substitutions can be automated for virtually any object language,
  whereas with intrinsic typing, it seems impossible to unify
  substitutions across syntactic categories, because an intrinsically
  typed multi-sorted syntax would require to index terms with terms,
  which is the same problem that occurs with giving intrinsic types to a dependently typed language.

  Our framework is able to automatically derive definitions and lemmas
  about untyped substitution, and provides an abstraction for proving
  type preservation of substitution with a single lemma (\ACode{\AField{\_⊢⋯\_}}).

  Similarly to frameworks with intrinsic typing, the main limitation in our work is the shape of
  typing relations: our typing kits can only model classical ternary typing relations.
  To adapt our framework to more complicated typing relations, the machinery for untyped substitution
  can be reused, but the typing kits and typing traversals need to be modified.
  We found that this works surprisingly well in practice,
  where we have already made extended typing kits to support
  linear typing ala Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} and
  general substructural typing ala Licata et al\cite{DBLP:conf/rta/LicataSR17}.
  In both cases, the typing relation has a fourth component which models usage restrictions
  on the type context.

>>>>

  Code of the framework is displayed in gray boxes.
  Code of examples is displayed in yellow boxes.
  Code of the System F formalization is displayed without boxes. The
  latter is the only code a user of our framework has to write.

  In this paper, we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in our supplementary material,
  which includes the simplified framework (368 lines of code) and the
  System F formalization (221 lines of code).
  The full framework including the case studies is available on Github:
  \textcolor{blue}{\url{https://github.com/m0rphism/kitty}}
  % The full framework is available online.\footnote{See supplementary material. In the final version, this text will be replaced by a link to github.}

@@

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  % \item
  %   an adaptation of McBride's
  %   kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  %   for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and its metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and type contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  We identify two drawbacks with this style of syntax:
  \begin{enumerate}
  \item
    the syntactic categories (\ACode{\ACon{Kind}},
    \ACode{\ACon{Type}}, and \ACode{\ACon{Expr}})
    have different types, which makes it difficult to treat them uniformly; and
  \item the different sorts of variables are modeled separately, which requires to define not just
    type-in-type and expression-in-expression substitution, but also type-in-expression substitution.
    Consequently, interaction lemmas between the substitutions are required.
    % In general, this leads to a combinatory explosion of
    % substitutions, e.g.\ adding kind variables would lead to a total of 6
    % different substitutions and corresponding interaction lemmas.
  \end{enumerate}
  To avoid these drawbacks, we instead use a multi-sorted syntax.

@@

  \begin{figure}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
    \label{fig:map-ops}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions.
  The definitions are included directly in the record module of the
  Kit, so they are implicitly parameterized over a kit.
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \ADef{\&} x} applies a map to a variable.
  The operation \ACode{ϕ \ADef{↑} s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{\ADef{⦅} x/t \ADef{⦆}} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\ADef{weaken}} map increases all variables by one.
  The \ACode{ϕ₁ \ADef{$\sim$} ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{\AField{$\sim$-ext}}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{\ADef{id}}.
  The lemma \ACode{\ADef{id↑-id}} states that a lifted identity map is again an identity map.

  To make it easier to talk about a specific kit, we introduce the following notations:
  % We also specify notation 
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  % We also specify notation, which makes it less cumbersome
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  \begin{itemize}
  \item
    we write \ACode{S₁ \ADef{–[} K \ADef{]→} S₂} for the \ACode{S₁ \ADef{→ₖ} S₂}
    of some specific \ACode{\ACon{Kit}} \ACode{K}.
  \item
    we write \ACode{S \ADef{∋/⊢[} K \ADef{]} s} for the \ACode{S ∋/⊢ s}
    of some specific \ACode{\ACon{Kit}} \ACode{K}; and
  \end{itemize}

  The operation of applying a map to a term depends on the concrete
  object language. It is captured by the following structure:
  \LibCode\KTraversal
  \noindent The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

@@

  First, we define the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}}. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using
    \AField{\_↑\_} to account for the bound variable before
    we apply it to the subterm.
  \end{itemize}

@@

  A compose kit \ACode{\ACon{CKit}~K₁~K₂~K₁⊔K₂}
  describes the operations necessary for definining the composition of a
  \ACode{K₁}-map with a \ACode{K₂}-map that results in a \ACode{K₁⊔K₂}-map:
  \LibCode\KCKit
  \noindent The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency\cite{DBLP:conf/esop/Jones00} and is determined
  by the choice of \ACode{K₁} and \ACode{K₂}.
  The fields of a compose kit have the following meaning:
  \begin{itemize}
  \item
    The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
    \ACode{x/t} (according to \ACode{K₁}) and a renaming or substitution
    \ACode{ϕ} (according to \ACode{K₂}) and applies \ACode{ϕ} to \ACode{x/t} resulting in
    a variable or term (according to \ACode{K₁⊔K₂}).
    From this operation we derive map composition \ACode{\ADef{\_·ₖ\_}} as
    shown in the previous subsection.
    % We use the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}, if we need to
    % make it explicit that we are using the composition of some
    % specific \ACode{\ACon{CKit}} \ACode{C}.
  \item 
    The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
    \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
    subsequent lemmas to make use of the lemmas that we have already proved
    for \ACode{\AField{\_⋯\_}}.
  \item 
    The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
    weakening is the same as weakening first and then lifting the map over
    the variable introduced by the weakening.
    From this lemma, we can derive that lifting distributes over composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}

@@

    % A weakening followed by a singleton substitution cancel each other out:
    % \LibCode\KWeakenCancelsSingle
    % % \LibCode\KWeakenCancelsSingleProof
    % By applying \ACode{\AField{fusion}} and \ACode{\AField{⋯-id}} to
    % this lemma, we get the same statement for multiple map applications:
    % \LibCode\KWeakenCancelsSingleTraverse
    % % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingleTraverse
    % A singleton map can be swapped with any map \ACode{ϕ}:
    % \LibCode\KDistLiftSingle
    % % \LibCode\KDistLiftSingleProof
    % By applying \ACode{\AField{fusion}} to this lemma, we get the
    % same statement for multiple map applications:
    % \LibCode\KDistLiftSingleTraverse
    % % \LibCode\KDistLiftSingleTraverseProof
  \end{itemize}
  Similarly, as it was the case for the \ACode{\ADef{Kit}} and \ACode{\ADef{Traversal}}
  structures, the idea is that we instantiate the \ACode{\ADef{CTraversal}} for
  our object language, and in return the framework defines the concrete \ACode{\ADef{CKit}}
  instances for us. Hence, we define the \ACode{\ADef{CKit}} instances in the record module of \ACode{\ADef{CTraversal}}:
  \LibCode\KCKitInstances
  \noindent
  \ACode{\ADef{Cᵣ}} is the compose kit between a renaming and another kit \ACode{K}.
  \ACode{\ADef{Cₛ}} is the compose kit between a substitution and another kit \ACode{K},
  and requires that we already know how to compose a \ACode{K}-map with a renaming.
  The following verifies that \ACode{\ADef{Cᵣ}} and \ACode{\ADef{Cₛ}} indeed get us all four compositions:

@@

  \AppCode\FPreserve
  The type of \ACode{\ADef{\_⊢⋯\_}} is the same as in the record definition and hence omitted.
  The interesting parts of the proof are:
  \begin{itemize}
  \item
    There is a strong similarity to the instantiation of
    map traversal \ACode{\ADef{\_⋯\_}}: where \ACode{\ADef{\_⋯\_}}
    used \ACode{\AField{`/id}} or \ACode{\AField{\_↑\_}}, our \ACode{\ADef{\_⊢⋯\_}} uses their
    preservation lemmas \ACode{\AField{⊢`/id}} or \ACode{\AField{\_⊢↑\_}}.
  \item
    The lambda typing constructor \ACode{\ACon{⊢λ}} weakens the type \ACode{t₂} to
    shield it from expression-substitution, requiring us to use
    \ACode{\AField{⋯-↑-wk}} to move the map under the weakening.
  \item
    The type application constructor \ACode{\ACon{⊢∙}} substitutes \ACode{t₁} into \ACode{t₂},
    requiring us to use \ACode{\AField{dist-↑-⦅⦆}} to move the map
    under the singleton substitution.
  \end{itemize}
  % Finally, we instantiate the \ACode{\ACon{TTraversal}} structure:
  % \AppCode\FTTraversal

@@

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    works out of the box for both deriving untyped substitution
    and instantiating the \ACode{\ACon{TTraversal}}.
    In the proof of confluence, the \ACode{\ACon{CKit}} abstraction
    allowed us to unify lemmas about the reduction of renamings and
    substitutions.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution-preserves-typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we instead used an encoding that allows us to use the
    the \ACode{\ACon{TKit}} abstraction directly.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accessed by the user.
    With this encoding, substitution-preserves-typing is generally true
    again: replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t}, results in a term that is still
    well-typed, but in a context with an unsatisfiable constraint \ACode{t' <: t}.
  \item
    % Object languages with pattern matching can be encoded by adding
    % the sorts of the variables bound by a pattern to the pattern sort
    % \ACode{\ACon{𝕡}} itself:
    % \begin{quote}
    % \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    % \end{quote}
    % This allows the clause of a \ACode{\ACon{match}} expression to
    % read the bound variables from the pattern sort and make them
    % available as free variables in its body expression.
    Object languages with pattern matching can be modeled by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself. A pattern matching clause \ACode{p~\ACon{⇒}~e} can
    then be defined as
    \begin{quote}
    \ACode{\ACon{\_⇒\_} \AKw{:} S \ACon{⊢} \ACon{𝕡} S' \AKw{→} \AKw{(}S' \ADef{++} S\AKw{)} \ACon{⊢} \ACon{𝕖} \AKw{→} S \ACon{⊢} \ACon{𝕔}} 
    \end{quote}
    where \ACode{S'} describes the variables bound by the pattern and \ACode{\ACon{𝕔}} is the sort of a clause.
  \end{itemize}

@@

  \subsection{Variable Binding}
  There is a plethora of different methods for representing
  variable binders:
  de Bruijn indices\cite{scholar:de1972lambda},
  co de Bruijn indices\cite{DBLP:journals/corr/abs-1807-04085},
  locally nameless\cite{DBLP:journals/jar/Chargueraud12},
  locally named\cite{DBLP:conf/tlca/McKinnaP93},
  higher order abstract syntax\cite{DBLP:conf/pldi/PfenningE88} and
  its parametric variant\cite{DBLP:conf/icfp/Chlipala08},
  nominal logic\cite{DBLP:journals/tcs/UrbanPG04},
  shifted names\cite{scholar:dolansyntax},
  nameless painless\cite{DBLP:conf/icfp/Pouillard11}, and
  scope graphs\cite{DBLP:journals/pacmpl/AntwerpenPRV18}.
  Many of these representations have been studied in solutions to the POPLMark
  challenges\cite{DBLP:conf/tphol/AydemirBFFPSVWWZ05, DBLP:journals/jfp/AbelAHPMSS19}.

@@

  With only a slight modification to their framework, we can instantiate it
  for multi-sorted syntax, enabling the definition of polymorphic languages.
  However, as the intrinsic typing is then
  used to describe syntactic categories, the semantic abstractions talk about untyped
  terms, so typing and type preservation lemmas have to be modeled entirely manually.
  We believe it would be worthwhile to explore how their semantics
  abstractions can be lifted to typing relations similar to how our typing kits lift
  regular kits from terms to typing relations.

@@

  We have presented an Agda framework, which automatically derives definitions and lemmas
  for untyped substitution, and provides an abstraction for proving
  type preservation of renaming and substitution for all syntactic
  categories with a single lemma (\ACode{\AField{\_⊢⋯\_}}).

  Compared to many extrinsically typed approaches, our framework also models
  typing and type preservation. Compared to many intrinsically typed approaches, our
  framework gracefully extends to polymorphic scenarios.

  The main limitation of our framework is the shape of typing relations, similarly
  as it is the case with approaches based on intrinsic typing: we can
  only model classical ternary typing relations.
  To adapt our framework to more complicated typing relations, the
  machinery for untyped substitution can be reused, but the
  abstractions related to typing need to be modified.
  We found that this works surprisingly well in practice,
  where we have already made extended typing abstractions that support
  linear typing ala Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} and
  general substructural typing ala Licata et al\cite{DBLP:conf/rta/LicataSR17}.
  In both cases, the typing relation is extended with a fourth
  component that models usage restrictions on the type context.

  While intrinsic typing allows to unify definitions with their type
  preservation proofs, extrinsic typing allows to unify
  substitutions across different syntactic categories, as we have demonstrated.
  We believe this makes our framework particularly suited for polymorphic languages,
  where the downside of extrinsic typing is automated away, and the benefits
  of a unified substitution bears fruits.
  % Furthermore, extrinsic typing allows us
  % to avoid the ``subst-hell'', which typically appears in formalizations
  % of logical relations for polymorphic languages.
  % We believe this makes extrinsic typing more suitable for polymorphic languages:
  % with extrinsic typing, the definitions and lemmas for untyped
  % substitutions can be automated for virtually any object language,
  % whereas with intrinsic typing, it seems impossible to unify
  % substitutions across syntactic categories, because an intrinsically
  % typed multi-sorted syntax would require to index terms with terms,
  % which is the same problem that occurs with giving intrinsic types to a dependently typed language.
